Optimizing Abstract$^{\text 2}$ Machines

### 2012-11-11 paper/paper.tex
Checkpoint intro and at a glance draft.

\includegraphics[width=2.5in]{rel-time-church.ps}

\subsection{Notation and prerequisites}

Prereqs: Semantics Engineering \cite{dvanhorn:Felleisen2009Semantics},
AAM \cite{dvanhorn:VanHorn2011Abstracting}
\cite{dvanhorn:VanHorn2012Systematic}.

@@

\newpage
\section{Starting point}
\subsection{An abstract$^2$ machine for ISWIM}

@@

\begin{align*}
\mathit{push}_\mlab^\mcntr(\msto,\mkont) &=
  \mlab\mcntr,\msto\sqcup[\mlab\mcntr\mapsto\{\mkont\}] \\
\mathit{bind}^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto \maddr],
                                           \msto\sqcup[\maddr \mapsto
                                             \{ \mval\}],
                                           \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= x\mcntr' \\
%%              \lfloor \mtcntr \rfloor_k &= \mtcntr \\
              \lfloor \mcntr \rfloor_0 &= \mtcntr \\
              \lfloor \mlab\mcntr \rfloor_{k+1} &= \mlab\lfloor \mcntr\rfloor_k
\end{align*}
%% Redundant
%% \paragraph 0CFA

%% \begin{align*}
%% \mathit{push}(\mlab,\msto,\mkont) &= \mlab,\msto\sqcup[\mlab\mapsto\{\mkont\}]
%% \\
%% \mathit{bind}(\menv,\msto,\mvar,\mval) &= \menv[\mvar\mapsto\mvar],\msto\sqcup[\mvar\mapsto\{\mval\}]
%% \end{align*}

@@

%% \begin{align*}
%% \widehat{\mathit{push}}(\maddr,\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
%% % \mbox{ where }\maddr \notin\msto
%% \\
%% \widehat{\mathit{bind}}(\msto,\mvar,\mval) &= \menv[\mvar\mapsto\maddr],\msto\sqcup[\maddr\mapsto\{\mval\}]
%% \end{align*}

@@

\subsection{An example: Analyzing Church number computations}

Figure~\ref{fig:church} gives a small example of computing with Church
numerals.  The example, while small, is informative: (1) it can be
written in almost any modern programming language\footnote{It is
  written here using Scheme notation and makes use of the shorthands:
%
\begin{alltt}
   (define (((f x) y) z) ...) \(\equiv\)

   (define f (\(\lambda\) (x) (\(\lambda\) (y) (\(\lambda\) (z) ...))))
\end{alltt}

\begin{alltt}
   \church{n} \(\equiv\) (\(\lambda\) (s) (\(\lambda\) (z) (s\(\subscript1\) (s\(\subscript2\) ... (s\(\subscript{n}\) z)))))
\end{alltt}
},
%
and (2) it stresses an analyzer's ability to deal with complicated
environment and control structure arising from the use of higher-order
functions to model arithmetic.

Our baseline analyzer takes NN:NN minutes to analyze this program
(without store widening it times out after 30 minutes).  This is on
par with the performance of Sergey,~et al.'s analyzer, which takes
NN:NN minutes.

\newpage
\subsection{Generic fixpoint calculator}

@@

\begin{align*}
lift(f)(X) &= \bigcup\{f(x) \mid x \in X\}
\end{align*}

Then $eval(e) = {\mathcal F}(lift(\machstep))(\{\ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}\})$.

@@

\subsection{Store widening}

@@

\subsection{Baseline evaluation}

@@

\section{Precision preserving recipe}

@@

% Ian's sections: generators, implementation and compound data
\section{Implementation}

@@

\section{Evaluation}

@@

\section{Related work}

@@

\section{Conclusion}

>>>>

%% \subsection{Notation and prerequisites}

%% Prereqs: Semantics Engineering \cite{dvanhorn:Felleisen2009Semantics},
%% AAM \cite{dvanhorn:VanHorn2011Abstracting}
%% \cite{dvanhorn:VanHorn2012Systematic}.

@@

\section{At a Glance}

In the following sections, we develop an analyzer for a core
functional language and follow a series of steps of abstraction and
optimization.  At each step we evaluate the implementation on an
example that computes with Church numerals, shown in
figure~\ref{fig:church}.
%% \footnote{It is
%%   written here using Scheme notation and makes use of the shorthands:
%% %
%% \begin{alltt}
%%    (define (((f x) y) z) ...) \(\equiv\)
%%
%%    (define f (\(\lambda\) (x) (\(\lambda\) (y) (\(\lambda\) (z) ...))))
%% \end{alltt}
%%
%% \begin{alltt}
%%    \church{n} \(\equiv\) (\(\lambda\) (s) (\(\lambda\) (z) (s\(\subscript1\) (s\(\subscript2\) ... (s\(\subscript{n}\) z)))))
%% \end{alltt}}
The example, while small, is informative:
\begin{enumerate}
\item it can be written in most modern programming languages, and
%
\item it stresses an analyzer's ability to deal with complicated
environment and control structure arising from the use of higher-order
functions to model arithmetic.
\end{enumerate}

We start, in section~\ref{sec:aam}, by developing an abstract
interpreter according to the abstracting abstract machines approach
and observe the resulting implementation cannot analyze the example in
under 30 minutes.  In section~\ref{sec:baseline}, we perform a further
abstraction by widening the store.  The resulting analyzer sacrifices
precision for speed and is able to analyze the example in about 1
minute.  Without this abstraction step the analysis is exponential due
to per-state store variance.  This step is described by Van Horn and
Might and is necessary to make even small examples feasible.  We
therefore take the widened interpreter as the baseline for our
evaluation.

In section~\ref{sec:opt} we give a series of simple abstractions and
implementation techniques that, in total, drop the analysis time to
only 138 milliseconds, a speedup of 475 times.
Figure~\ref{fig:churchtime} shows the step-wise improvement of the
analysis time for this example.

\begin{figure}
\begin{center}
\includegraphics[width=2.5in]{rel-time-church.ps}
\end{center}
\caption{Relative run-time of
optimized analyers to baseline for Church numeral example.}
\label{fig:churchtime}
\end{figure}

In section~\ref{sec:impl}, we describe a more complete implementation
scaled up and capable of analyzing Scheme benchmark programs.

In section~\ref{sec:eval} we evaluate this implementation against a
set of benchmark programs drawn from the literature.  Many benchmarks
cause the baseline analyzer to take longer than 30 minutes or to
consume more 1 gigabit of memory, at which point the analysis is
stopped.  This is the case for the largest benchmark program, which is
3,500 lines of code and takes under a minute in the most optimized
analyer.  For those benchmarks that did complete on the baseline, the
optimized analyzer outperformed the baseline by a factor of
% 475
two to
% 4,382
three orders of magnitude.

Section~\ref{sec:related} relates this work to the literature and
section~\ref{sec:conclusion} concludes.

\section{An abstract$^{\text 2}$ machine for ISWIM}
\label{sec:aam}

@@

\begin{align*}
\mathit{push}_\mlab^\mcntr(\msto,\mkont) &=
  \mlab\mcntr,\msto\sqcup[\mlab\mcntr\mapsto\{\mkont\}] \\
\mathit{bind}^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto \maddr],
                                           \msto\sqcup[\maddr \mapsto
                                             \{ \mval\}],
                                           \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= x\mcntr' \\
%%              \lfloor \mtcntr \rfloor_k &= \mtcntr \\
              \lfloor \mcntr \rfloor_0 &= \mtcntr \\
              \lfloor \mlab\mcntr \rfloor_{k+1} &= \mlab\lfloor \mcntr\rfloor_k
\end{align*}

@@

@@

\newpage
\section{Baseline analyzer}
\label{sec:baseline}

\subsection{Generic fixpoint calculator}
\label{sec:fixpoint}

@@

\begin{align*}
\lift(f)(X) &= \bigcup\{f(x) \mid x \in X\}
\end{align*}

Then $\eval(e) = {\mathcal F}(\lift(\machstep))(\{\ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}\})$.

@@

\subsection{Store widening}
\label{sec:storewiden}

@@

\subsection{Baseline evaluation}
\label{sec:baselineeval}

@@

\section{Precision preserving recipe}
\label{sec:opt}

@@

% Ian's sections: generators, implementation and compound data
\section{Implementation}
\label{sec:impl}

@@

\section{Evaluation}
\label{sec:eval}

@@

\section{Related work}
\label{sec:related}

@@

\section{Conclusion}
\label{sec:conclusion}
